---
title: "Project_CS1380_PN"
author: "Phuong Nguyen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load + Clean data

The data set "heart_failure" features 299 patients with their conditions and status related to Heart Failure in 2005.

```{r}
heart <- read.csv("heart_failure.csv", header = TRUE)
head(heart)

ind <- which(is.na(heart))
ind

heart$sex <- factor(heart$sex, levels = c(0,1), labels = c("Male","Female"))
heart$DEATH_EVENT <- factor(heart$DEATH_EVENT, levels = c(0,1), labels = c("No","Yes"))
heart$high_blood_pressure <- factor(heart$high_blood_pressure, levels = c(0,1), labels = c("No High Blood Pressure","High Blood Pressure"))
heart$diabetes <- factor(heart$diabetes, levels = c(0,1), labels = c("No Diabetes", "Diabetes"))

```

The objective of this project is to identify a pattern among patients suffering from Heart Failure, and to eventually develop a predictive model for mortality by using specific underlying factors.


## Extract relevant data + visualization

-   Examine: Age, Have high blood pressure, Have diabetes, Level of creatinine_phosphokinase (indicator of heart failure), Ejection fraction (percentage of blood leaving the heart per contraction), Gender and Mortality
-   For Age: Usually heart-related illnesses happen more frequently in elderly age group. However, in recent studies, it is suggested that the age range for having heart conditions is shifting "younger". From the original value, if the age is under 45, it is considered "young"; between 45 and 65 is "middle-age", and over 65 is consider "senior".

```{r, warning = FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
new_heart <- heart %>%
  select(c(age,high_blood_pressure,diabetes,creatinine_phosphokinase, ejection_fraction, sex, DEATH_EVENT)) %>%
  mutate(Age_range = case_when(age <= 45 ~ "Young",
                               age <= 65 ~ "Middle-age", 
                               age > 65 ~ "Senior")) %>%
  rename(CPK_level = creatinine_phosphokinase, HBP = high_blood_pressure, Ejection_percentage = ejection_fraction, Diabetes = diabetes, Mortality = DEATH_EVENT, Gender = sex, Age = age)
head(new_heart)
```

## Distribution of patients' Age in the study

We start by examining the distribution of patients' age in this data set

```{r}
ggplot(new_heart, aes(x = Age, fill = Age_range)) +
  geom_histogram(binwidth = 1, colour = "white") +
  ylab("Number of patients\n") +
                ggtitle("               Distribution of Patient's Age\n")
```

The histogram is right-skewed. The majority of patients with heart failures are middle-age and senior.

## Visualise number of paients who have high blood pressure or diabetes and the association to Mortality

Next, underlying health conditions are compared against mortality to see if there is an association.

```{r}
ggplot(new_heart, aes(x = Mortality, fill = Diabetes)) +
  geom_bar(position = "dodge") +
  facet_grid(.~ HBP) + ylab("Number of patients\n") +
  ggtitle("        Association between Mortality by Heart Failures 
                        and Underlying Health Conditions\n")

```

In general, from the bar plot, having diabetes or high-blood-pressure do not contribute significantly to a higher chance of mortality.

## Visualize the relationship between Mortality and CPK level in blood

Creatinine phosphokinase (CPK) is an enzyme that is found mainly in muscle tissues. The normal level of CPK in blood usually ranges from 10 -120 mcg/L, and when there is a muscle injury or inflammation, CPK is released causing a surge. Here, the relationship between CPK_level and Mortality is visualized to see if there is a positive association.

```{r}
library(ggplot2)
ggplot(new_heart, aes(x = Mortality, y = CPK_level)) +
  geom_boxplot() + 
  ggtitle("Relationship between Mortality and Creatinine Phosphokinase 
                    level of patients with heart injury\n") +
  ylab("Creatinine Phosphokinase level (mcg/L)\n")

```




The boxplot shows that in general, there is little difference in the mean of CPK level in blood between patients who survive or die until the next health check-up. A possible explanation for the insignificant difference is because all patients in this study have suffered from heart failures in varied degrees, the CPK values are high already for all patients. There are, however, many outliers of CPK level in the upper region for both groups.

## Visualize the relationship between Mortality and Percentage of blood leaving the heart per contraction (Ejection_percentage/EP)

Ejection percentage (EP) is the percentage of blood leaving the heart per contraction. A normal ejection percentage is between 55% and 75%; the range 40 - 55% is considered low function, and under 40% means there is a risk for possible heart failure. The values of Ejection_percentage in this data set are visualized and compared to Mortality to see if there is any relationship.

```{r}
library(ggplot2)
ggplot(new_heart, aes(x = Mortality, y = Ejection_percentage)) +
  geom_boxplot() + 
  ggtitle("Relationship between Mortality and the blood ejection percentage 
                 from the heart of patients with heart injury\n") +
  ylab("Ejection percentage (%)\n")

```



The boxplot shows that between survived and non-survived patients, the average percentage of blood leaving the heart per contraction is lower than that in survived patients. It suggests that with less blood flow from the heart, the risk of mortality is higher. In the group of non-survived patients, there are some outliers in the 60% and above range. This result agrees with medical knowledge that too high of ejection percentage is caused by some issues such as hypertrophic cardiomyopathy, which leads to sudden cardiac arrest.

## Summarize of Average CPK, Average EP and Mortality rate by Gender and Age_group

Different genders and Age groups can have an effect to a certain extent on mortality.

```{r}
new_heart %>%
  group_by(Gender, Age_range) %>%
  summarise(Average_CPK = mean(CPK_level), Average_EP = mean(Ejection_percentage), Mortality_Rate = mean(Mortality == "Yes"))
```

-   From the summary table for Mortality rate among genders, age ranges, average CPK levels and average Ejection percentage, in this study, Senior age group has the highest mortality rate for both males and females, followed by middle-age group and young group.
-   However, for females, the mortality rates are equal between the young and middle-age group, with the young group having a significantly higher average CPK level than the middle-age.
-   Between males and females, ejection percentage, on average, is higher in males than in females.

## Prediction model

There are several types of the CPK enzyme that are associated with different organs. When a person undergoes a heart attack or heart injury, the CPK-2 level is elevated in the blood. Hence, CPK level in the blood can be used as an indicator for heart attack. Additionally, Ejection percentage (EP) is believed to be a contributor to the mortality of patients in heart failure. This project aims to use CPK and EP to predict the mortality rate. The predicted model will then be compared with actual data by using a confusion matrix to see how much accuracy it can achieved.

```{r, warning=FALSE, message=FALSE}
#model 1
pred.model <- glm(Mortality ~ CPK_level + Ejection_percentage, new_heart, 
                  family = "binomial")
summary(pred.model)

#model 2
pred.cpk <- glm(Mortality ~ CPK_level, data = new_heart, family = "binomial")
summary(pred.cpk)

#model 3
pred.ep <- glm(Mortality ~ Ejection_percentage, data = new_heart, family = "binomial")
summary(pred.ep)

```

## Analysis of prediction models 1 & 3

- Three logistic regression model have a common result that the p-value of ejection percentage is statistically significant (<0.001), while the p-value of CPK-level in the all models is statistically insignificant (>0.1). Hence, model 3 and possibly model 1 should be used for prediction of mortality. 
- In both model 1 and model 3: When 'Ejection_percentage' increases by 1 unit, the log-odd of patients' mortality decreases by ~0.056 unit
- Below is the prediction of mortality when using model 1 and 3 and a standard cutoff value at 0.5

```{r}
#Model 1
pred_mort_1 <- predict(pred.model, new_heart, type = "response")
pred_lab_1 = if_else(pred_mort_1 > 0.5, "Yes","No")
pred_lab_1 = factor(pred_lab_1, labels = c("No","Yes"))
cfs.mtx.1 = table(pred_lab_1, new_heart$Mortality)
cfs.mtx.1

accuracy.1 = (sum(diag(cfs.mtx.1)))/sum(cfs.mtx.1)
accuracy.1

sensitivity.1 = cfs.mtx.1[2,2]/sum(cfs.mtx.1[,2])
sensitivity.1

specificity.1 = cfs.mtx.1[1,1]/sum(cfs.mtx.1[,1])
specificity.1



#Model 3
pred_mort_3 <- predict(pred.ep, new_heart, type = "response")
pred_lab_3 = if_else(pred_mort_3 > 0.5, "Yes","No")
pred_lab_3 = factor(pred_lab_3, labels = c("No","Yes"))
cfs.mtx.3 = table(pred_lab_3, new_heart$Mortality)
cfs.mtx.3

accuracy.3 = (sum(diag(cfs.mtx.3)))/sum(cfs.mtx.3)
accuracy.3

sensitivity.3 = cfs.mtx.3[2,2]/sum(cfs.mtx.3[,2])
sensitivity.3

specificity.3 = cfs.mtx.3[1,1]/sum(cfs.mtx.3[,1])
specificity.3
```

Model 1 has an accuracy of 0.75, a sensitivity of 0.25 and a specificity of 0.98. Model 3 has an accuracy of 0.74, a sensitivity of 0.21 and a specificity of 0.99. A high specificity means the models correctly identify the true negatives, aka the number of survived patients (no mortality). However, the two models have low sensitivity, which means they cannot correctly predict the true positives, aka the number of non-survived patients (mortality). 

## Examine accuracy by choosing cutoff values

A high cut-off value is prone to produce more false negative (less mortality than actual cases), while a low cut-off value is prone to produce more false positive (more mortality than actual cases). A range of cutoff value is chosen to evaluate the accuracy of model 1 and 3

```{r}

cutoff_seq = seq(from = 0.1, to = 0.8, by = 0.05)

#Model 1
accuracy.1.seq = character(length = length(cutoff_seq))
pred_mort_1 <- predict(pred.model, new_heart, type = "response")
for (i in 1:length(cutoff_seq)){
  cutoff = cutoff_seq[i]
  pred_lab_1 = if_else(pred_mort_1 > cutoff, "Yes","No")
  #pred_lab_1 = factor(pred_lab_1, labels = c("No","Yes"))
  cfs.mtx.1 = table(pred_lab_1, new_heart$Mortality)
  accuracy_1 = (sum(diag(cfs.mtx.1)))/sum(cfs.mtx.1)
  accuracy.1.seq[i] = round(accuracy_1,2)
}

#Model 3
accuracy.3.seq = character(length = length(cutoff_seq))
pred_mort_3 <- predict(pred.ep, new_heart, type = "response")
for (i in 1:length(cutoff_seq)){
  cutoff = cutoff_seq[i]
  pred_lab_3 = if_else(pred_mort_3 > cutoff, "Yes","No")
  #pred_lab_1 = factor(pred_lab_1, labels = c("No","Yes"))
  cfs.mtx.3 = table(pred_lab_3, new_heart$Mortality)
  accuracy_3 = (sum(diag(cfs.mtx.3)))/sum(cfs.mtx.3)
  accuracy.3.seq[i] = round(accuracy_3,2)
}

df = data.frame(Cutoff_Values = cutoff_seq, Accuracy.1 = accuracy.1.seq, Accuracy.3 = accuracy.3.seq)

ggplot(df, aes(x = cutoff_seq)) +                 
  geom_point(aes(y= accuracy.1.seq, color = "Model 1"), pch = 19, size = 1) +
  geom_point(aes(y=accuracy.3.seq, color = "Model 3"), pch = 19,alpha = 0.3, size = 2) +
  labs(x = "\nCutoff values", y = "Accuracy\n", 
       title = "  Accuracy of Model 1 & Model 3 based on cutoff values \n", color = "Models") +
  scale_color_manual(values = c("Model 1" = "blue", "Model 3" = "red"))



```

From the plot of accuracy from two predictive models, it can be observed that the maximum of achievable accuracy is around ~0.75, when the cutoff is at 0.5. Although the accuracy can be classified as "relatively well", more factors should be considered in the logistic regression model or a different type of prediction models should be used to optimize the procedure of heart-injury diagnosis.


#### References

[1] Larxel. (2020, June 20). Heart failure prediction. Kaggle. https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data

[2] Chicco, D., Jurman, G. Machine learning can predict survival of patients with heart failure from serum creatinine and ejection fraction alone. BMC Med Inform Decis Mak 20, 16 (2020). https://doi.org/10.1186/s12911-020-1023-5

[3] Creatine phosphokinase test. Mount Sinai Health System. (n.d.). https://www.mountsinai.org/health-library/tests/creatine-phosphokinase-test#:~:text=Normal%20Results,per%20liter%20(mcg%2FL)

[4] Ejection Fraction: What the Numbers Mean. Pennmedicine.org. (2022, April 13). https://www.pennmedicine.org/updates/blogs/heart-and-vascular-blog/2022/april/ejection-fraction-what-the-numbers-mean#:~:text=Ejection%20fraction%20is%20measured%20as,blood%20and%20may%20be%20failing. 
